Inference for multivariate peaks over threshold models

### 2021-10-12 introduction.tex
updates
update

\section{Introduction}
The statistical analysis of extreme values focuses on inference for
rare events that correspond to the tails of probability distributions.
As such, it is a key ingredient in the assessment of the risk of
phenomena that can have strong societal impacts like floods, heat waves,
high concentration of pollutants, crashes in the financial markets,
among others. The fundamental challenge of extreme value theory (EVT) is
to use information, collected over limited periods of time, to
extrapolate to long time horizons. This sets EVT apart from most of
statistical inference, where the focus is on the bulk of the
distribution. Extrapolation to the tails of the distributions is
possible thanks to theoretical results that give asymptotic descriptions
of the probability distributions of extreme values.

@@

  The reminder of this paper is outlined as follows. {\bf We need a roadmap. Part of the roadmap should mention the analysis of the IVT. Basically condensing the paragraph below in a couple of sentences.}

To motivate the topic and our proposed solution, we acknowledge the relevance of extreme analysis to
  climatological events.  Outcomes such as heat waves, extreme precipitation, storm surges, and
  such present a similar challenge to investigators, in that the events of interest can be very destructive,
  and occur in probability at the tail end of \emph{normal} behavior.  In this fashion, a model built
  to capture \emph{normal} behavior may struggle to adequately represent extreme behavior.  More
  specialized tools become necessary; especially so in a time of intensifying extreme weather
  events~\citep{jentsch2007,vousdoukas2018,li2019}.  To this end, we use the integrated water vapor
  transport (\emph{IVT}) data for observing atmospheric rivers as an example application for our proposed method.

@@

\section{A multivariate PoT model}
%\subsubsection{Univariate Maxima}
%Regarding the asymptotic behavior of extreme events, there are a couple major strategies for conducting
%  inference---developing probabilistic estimates of extreme behavior.  First developed is the theory
%  of a limiting distribution on maxima---the largest observation from a sample.  For a sample $\bm{x}$
%  where $\bm{x} = (x_1,\ldots,x_n)$ represents a sequence of $n$ independent random variables from a
%  distribution function $F$, the distribution of the maximum $M_n$ of this sequence can be derived
%  as:
%  \begin{equation*}
%    \begin{aligned}
%      \text{Pr}\left[M_n\leq z\right] &= \text{Pr}\left[X_1 \leq z, \ldots, X_n \leq z\right]\\
%        &= \text{Pr}\left[X_1\leq z\right]\times\ldots\times\text{Pr}\left[X_n\leq z\right]\\
%        &= F(z)^n.
%    \end{aligned}
%  \end{equation*}
%  In situations where $F$ is unknown, we can seek to approximate the behavior of $F^n$ as
%  $n\rightarrow\infty$.  To ensure this does not degenerate to a point mass, we need to consider a
%  standardized sequence of maxima. If this sequence stabilizes as $n$ increases, the a limiting
%  distribution exists.  More specifically, if there exists some sequence of constants $a_n > 0$, $b_n$
%  such that:
%  \begin{equation*}
%    \text{Pr}\left[\frac{M_n - b_n}{a_n} \leq z\right] \stackrel{d}{\rightarrow} G(z)
%  \end{equation*}
%  as $n\rightarrow\infty$, then $G$ is a max-stable distribution, and $F$ is in the domain of
%  attraction of that max stable distribution.  \cite{frechet1927} originates the
%  field, identifying two limiting forms of max-stable distributions, which would become known as the
%  Fr{\'e}chet and Weibull distributions. \cite{weibull1951} expands the analysis of the Weibull
%  distribution.  \cite{fisher1928}, in what is now called the \emph{Fisher-Tippett Theorem}, identifies
%  the Fr{\'e}chet and Weibull distributions, along with an as then unnamed third form, as the three
%  limiting forms of the distribution of the maxima of a sample.  \cite{gumbel1935,gumbel1942} offer
%  analysis of that third form, now known as the Gumbel distribution.  Later works, including
%  \cite{jenkinson1955} reparameterize all three forms as special cases of a single unifying form,
%  the generalized extreme value distribution, \emph{GEV}:
%  \begin{equation*}
%    \label{eqn:gev}
%    F(m \mid \mu, \sigma, \xi) = \exp\left\lbrace-\left[1 + \xi\left(\frac{x - \mu}{\sigma}\right)\right]_{}^{-1/{\xi}}\right\rbrace.
%  \end{equation*}
%  Thus distributions in the domain of attraction of an EVD (Gumbel, Fr{\'e}chet, or Weibull) will be
%  in the domain of attraction of the GEV.

@@

To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty)^d$, the positive $d$-dimensional cone. Following the discussion in \cite{goix2017} we assume a form of regular variation, consisting on assuming the existence of a limiting measure $\mu$, such that.
\[
    \lim\limits_{n\to\infty}n\text{Pr}\left[\frac{1}{n}\bm{Z}\geq \bm{z}\right] = \mu\left([\bm{0},\bm{z}]^c\right),
\]
where the inequality and the interval are taken componentwise. 
In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions, and it is referred to as the
  \emph{exponent measure}. Under this condition $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
Define that $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
These two assumption together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
\cite{ferreira2014}, applied to the finite-dimensional case.
Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Moreover,
for some measure $\Phi$, $\mu(\cdot)$ is factorized as
\begin{equation}\label{mufact}
    \mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
\end{equation}
$R$ describes the radial component, and $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$ the 
angular component of $\bm{Z}$.  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
From Equation (\ref{mufact}) we see that $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
\begin{equation} \label{MGP}
    \text{Pr}\left[\bm{V} \in A \mid R > r\right]
      = \frac{r\text{Pr}\left[\bm{V} \in A, R > r\right]}{r\text{Pr}[R > r]}\hspace{0.2cm}
      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi(\mathcal{S}_{\infty}^{d-1})}.
\end{equation}
 Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $R$ being large.
 As such, it controls the dependence structure of $\bm{Z}$ in the tails. In fact, following \cite{ferreira2014} the distribution function of $\bm{Z}$ is given as
\begin{equation*}
	\text{Pr}[Z_1\leq z_1, \ldots ,Z_d\leq z_d] = E\left(\bigvee_{i=1}^d
\frac{V_i}{z_i\wedge 1}\right) - E\left(\bigvee_{i=1}^d
\frac{V_i}{z_i}\right) 
\end{equation*}
where the expectation is taken with respect to $\Phi$.
In view of this, to obtain a PoT model for $\bm{Z}$, we seek 
a flexible distribution on $\bm{V} \in \mathcal{S}_{\infty}^{d-1}$.

@@

A measure that is used to characterize the strength of the dependence,
in the tail, for two random variables $W_1$ and $W_2$, with marginal
distributions $F_1$ and $F_2$ is given by
\[	\chi_{12} = \lim_{u\uparrow 1} \text{Pr}[ F_1(W_1)>u|F_2(W_2)>u]  \]
\citep{coles2001}.
$\chi_{12}$ provides information about the distribution of extremes for the variable $W_1$
given that $W_2$ is very large.  When $\chi_{12}>0$, $W_1$ and $W_2$ 
are said to
be asymptotically dependent, otherwise they are asymptotically
independent. The following result provides the asymptotic dependence coefficient for two components of $\bm{Z}$. 
\begin{prop} \label{ppchi}
Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
$\text{Pr}[V_i>0] = 1$ and $E(V_i)$ exists ,
$i=1, \ldots ,d$, then
\begin{equation}
	\chi_{ij} = E\left(\frac{V_i}{E(V_i)} \wedge 
\frac{V_j}{E(V_j)}\right).
\end{equation}
\end{prop}
{\em Proof:}
Denote as $F_i$ the marginal distribution of $Z_i$. To obtain $\chi_{ij}$ we need $Pr(Z_1>z_1,Z_2>z_2)$, where $z_i =
F_i^{-1}(u) = E(V_i)/(1 - u), \;i=1,\dots, d$.
Using the fact that
$V_i>0, \forall i$ almost surely, we have that the former is equal to
\[	\text{Pr}\left[R>\frac{z_i}{V_i}\vee\frac{z_2}{V_j}\right] =
E\left(1\wedge\left(\frac{z_i}{V_i}\vee\frac{z_j}{V_j}\right)
^{-1} \right)
= E\left( \frac{V_i}{z_i}\wedge\frac{V_j}{z_j}
\right) = (1 - u)E\left(
\frac{V_i}{EV_i}\wedge\frac{V_j}{EV_j}
\right) ,
\]
where the second identity is justified by the fact that $V_i$ is
bounded and $z_i\rightarrow\infty$. The proof is completed by noting
that $Pr(F_i(Z_i)>u ) = 1 - u.\hfill\Box$

For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function. The following result provides the relevant expression, as a function of the angular measure.
\begin{prop}
Assume the same conditions of Proposition \ref{ppchi}. 
Let $\alpha \subset \{1, \ldots ,p\}$ be a collections of indexes. 
Then     
\begin{equation} \label{eqn:condsurv2df}
    \text{Pr}\left[\bigcap_{l \in \alpha} Z_l > z_l \right|\left. \bigcap_{l\not\in\alpha} Z_l > z_l\right] =
      \frac{E\left(\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}\right)}{E\left(\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}\right)}.
  \end{equation}
\end{prop}  
The proof uses a similar approach to the proof of Proposition \ref{ppchi}.

>>>>

\section{Introduction}

The statistical analysis of extreme values focuses on inference for
rare events that correspond to the tails of probability distributions.
As such, it is a key ingredient in the assessment of the risk of
phenomena that can have strong societal impacts like floods, heat waves,
high concentration of pollutants, crashes in the financial markets,
among others. The fundamental challenge of extreme value theory (EVT) is
to use information, collected over limited periods of time, to
extrapolate to long time horizons. This sets EVT apart from most of
statistical inference, where the focus is on the bulk of the
distribution. Extrapolation to the tails of the distributions is
possible thanks to theoretical results that give asymptotic descriptions
of the probability distributions of extreme values.

@@

  The reminder of this paper is outlined as follows. Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, detailing the separation of the radial measure from the angular measure.
  Section~\ref{sec:methodology} details our routine for estimating the angular measure, where~\ref{subsec:projection} elaborates the process of projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ onto an a series of manifolds ${\mathbb S}_p^{d-1}$, $p = 1,2,\ldots$ approaching the support of the angular measure, on to which~\ref{subsec:projgamma} builds to develop a \emph{projected Gamma} family of distributions.  Section~\ref{sec:evaluation} is concerned with developing criteria for model selection in the support of the angular measure.  Section~\ref{sec:results} details the efficacy of the projected gamma model in recovering the distribution of data on the support of the angular measure---in~\ref{subsec:simulated}, this is done on simulated data, and in~\ref{subsec:ivt}, acknowledging the relevance of of extreme value theory to climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, we estimate the extremal dependence structure for the integrated vapor transport---a measure of water vapor in the atmosphere, used for identifying atmospheric rivers.  Inferences from the dependence structure detailed in~\ref{subsec:ped} and~\ref{subsec:condsurv} are applied.  Finally, Section~\ref{sec:conclusion} concludes.

% Extreme value theory bears strong relevance to climate analysis, as climatological events of interest occur at the tail of the distribution, both in terms of magnitude and frequency.  We motivate our proposed model with an analysis of the integrated vapor transport data for observing atmospheric rivers---high concentrations of water vapor that lead to extreme precipitation events.

% To motivate the topic and our proposed solution, we acknowledge the relevance of extreme analysis to
%   climatological events.  Outcomes such as heat waves, extreme precipitation, storm surges, and
%   such present a similar challenge to investigators, in that the events of interest can be very destructive,
%   and occur in probability at the tail end of \emph{normal} behavior.  In this fashion, a model built
%   to capture \emph{normal} behavior may struggle to adequately represent extreme behavior.  More
%   specialized tools become necessary; especially so in a time of intensifying extreme weather
%   events.  To this end, we use the integrated water vapor
%   transport (\emph{IVT}) data for observing atmospheric rivers as an example application for our proposed method.

@@

\section{A multivariate PoT model\label{sec:multivariatepot}}
%\subsubsection{Univariate Maxima}
%Regarding the asymptotic behavior of extreme events, there are a couple major strategies for conducting
%  inference---developing probabilistic estimates of extreme behavior.  First developed is the theory
%  of a limiting distribution on maxima---the largest observation from a sample.  For a sample $\bm{x}$
%  where $\bm{x} = (x_1,\ldots,x_n)$ represents a sequence of $n$ independent random variables from a
%  distribution function $F$, the distribution of the maximum $M_n$ of this sequence can be derived
%  as:
%  \begin{equation*}
%    \begin{aligned}
%      \text{Pr}\left[M_n\leq z\right] &= \text{Pr}\left[X_1 \leq z, \ldots, X_n \leq z\right]\\
%        &= \text{Pr}\left[X_1\leq z\right]\times\ldots\times\text{Pr}\left[X_n\leq z\right]\\
%        &= F(z)^n.
%    \end{aligned}
%  \end{equation*}
%  In situations where $F$ is unknown, we can seek to approximate the behavior of $F^n$ as
%  $n\rightarrow\infty$.  To ensure this does not degenerate to a point mass, we need to consider a
%  standardized sequence of maxima. If this sequence stabilizes as $n$ increases, the a limiting
%  distribution exists.  More specifically, if there exists some sequence of constants $a_n > 0$, $b_n$
%  such that:
%  \begin{equation*}
%    \text{Pr}\left[\frac{M_n - b_n}{a_n} \leq z\right] \stackrel{d}{\rightarrow} G(z)
%  \end{equation*}
%  as $n\rightarrow\infty$, then $G$ is a max-stable distribution, and $F$ is in the domain of
%  attraction of that max stable distribution.  \cite{frechet1927} originates the
%  field, identifying two limiting forms of max-stable distributions, which would become known as the
%  Fr{\'e}chet and Weibull distributions. \cite{weibull1951} expands the analysis of the Weibull
%  distribution.  \cite{fisher1928}, in what is now called the \emph{Fisher-Tippett Theorem}, identifies
%  the Fr{\'e}chet and Weibull distributions, along with an as then unnamed third form, as the three
%  limiting forms of the distribution of the maxima of a sample.  \cite{gumbel1935,gumbel1942} offer
%  analysis of that third form, now known as the Gumbel distribution.  Later works, including
%  \cite{jenkinson1955} reparameterize all three forms as special cases of a single unifying form,
%  the generalized extreme value distribution, \emph{GEV}:
%  \begin{equation*}
%    \label{eqn:gev}
%    F(m \mid \mu, \sigma, \xi) = \exp\left\lbrace-\left[1 + \xi\left(\frac{x - \mu}{\sigma}\right)\right]_{}^{-1/{\xi}}\right\rbrace.
%  \end{equation*}
%  Thus distributions in the domain of attraction of an EVD (Gumbel, Fr{\'e}chet, or Weibull) will be
%  in the domain of attraction of the GEV.

@@

To develop a multivariate PoT model for extreme values consider a $d$-dimensional random vector $\bm{Z} \in {\mathbb R}^d_+ = [0,+\infty)^d$, the positive $d$-dimensional cone. Following the discussion in \cite{goix2017} we assume a form of regular variation, consisting on assuming the existence of a limiting measure $\mu$, such that.
\[
    \lim\limits_{n\to\infty}n\prob{\frac{1}{n}\bm{Z}\geq \bm{z}} = \mu\left([\bm{0},\bm{z}]^c\right),
\]
where the inequality and the interval are taken componentwise. 
In this setting, $\mu$ is the asymptotic distribution of $\bm{Z}$ in extreme regions, and it is referred to as the
  \emph{exponent measure}. Under this condition $\mu$ features the homogeneity property $\mu(tA) = \frac{1}{t}\mu(A)$. 
Define  $R = \max_{i} Z_i$, and assume that $R>1$ with probability one. Also, define $V_i = Z_i/\max_i Z_i$, and assume that $E(V_i)>0$.
These two assumption together with the homogeneity of $\mu$ correspond to the second condition in Theorem 2.1 of
\cite{ferreira2014}, applied to the finite-dimensional case.
Thus, we have the factorization $\bm{Z} =  R\bm{V}$, where $R$ and $\bm{V}$  are independent. Moreover,
for some measure $\Phi$, $\mu(\cdot)$ is factorized as
\begin{equation}\label{mufact}
    \mu\left( \left\lbrace\bm{Z} : V \in A , R > r\right\rbrace \right) = r^{-1}\Phi(A).
\end{equation}
$R$ describes the radial component, and $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$ the 
angular component of $\bm{Z}$.  Here ${\mathbb S}_\infty^{d-1}$ denotes the positive
orthant of the unit sphere in infinite norm.  $\Phi(\cdot)$ is referred to as the \emph{angular measure}.   
From Equation (\ref{mufact}) we see that $R$ is distributed as a standard Pareto random variable. Conditioning on $R$, we have that
\begin{equation} \label{MGP}
    \prob{\bm{V} \in A \mid R > r} = \frac{r\prob{\bm{V} \in A, R > r}}{r\prob{R > r}}\hspace{0.2cm}
      \xrightarrow[r\to\infty]{~} \hspace{0.2cm} \frac{\Phi(A)}{\Phi({\mathbb S}_{\infty}^{d-1})}.
\end{equation}
 Thus, the angular measure corresponds to the distribution of $\bm{V}$, conditional on $R$ being large.
 As such, it controls the dependence structure of $\bm{Z}$ in the tails. In fact, following \cite{ferreira2014} the distribution function of $\bm{Z}$ is given as
\begin{equation*}
    \prob{Z_1\leq z_1, \ldots ,Z_d\leq z_d} = \expect{\bigvee_{i=1}^d
\frac{V_i}{z_i\wedge 1}} - \expect{\bigvee_{i=1}^d
\frac{V_i}{z_i}}
\end{equation*}
where the expectation is taken with respect to $\Phi$.
In view of this, to obtain a PoT model for $\bm{Z}$, we seek 
a flexible distribution on $\bm{V} \in {\mathbb S}_{\infty}^{d-1}$.

@@

\subsection{Pairwise Extremal Dependence\label{subsec:ped}}
A measure that is used to characterize the strength of the dependence,
in the tail, for two random variables $W_1$ and $W_2$, with marginal
distributions $F_1$ and $F_2$ is given by
\[	\chi_{12} = \lim_{u\uparrow 1} \prob{F_1(W_1)>u|F_2(W_2)>u}  \]
\citep{coles2001}.
$\chi_{12}$ provides information about the distribution of extremes for the variable $W_1$
given that $W_2$ is very large.  When $\chi_{12}>0$, $W_1$ and $W_2$ 
are said to
be asymptotically dependent, otherwise they are asymptotically
independent. The following result provides the asymptotic dependence coefficient for two components of $\bm{Z}$. 
\begin{prop}\label{ppchi}
Suppose that $\bm{Z} = R\bm{V}$ with $R\sim Pa(1)$,
$\prob{V_i > 0} = 1$ and $\expect{V_i}$ exists ,
$i=1, \ldots ,d$, then
\begin{equation}
	\chi_{ij} = \expect{\frac{V_i}{\expect{V_i}} \wedge \frac{V_j}{\expect{V_j}}}
\end{equation}
\end{prop}
{\em Proof:}
Denote as $F_i$ the marginal distribution of $Z_i$. To obtain $\chi_{ij}$ we need $Pr(Z_1>z_1,Z_2>z_2)$, where $z_i =
F_i^{-1}(u) = \expect{V_i}/(1 - u), \;i=1,\dots, d$.
Using the fact that
$V_i>0, \forall i$ almost surely, we have that the former is equal to
\begin{equation*}
\prob{R>\frac{z_i}{V_i}\vee\frac{z_2}{V_j}}
    = \expect{1\wedge\left(\frac{z_i}{V_i}\vee\frac{z_j}{V_j}\right)^{-1}}
    = \expect{\frac{V_i}{z_i}\wedge\frac{V_j}{z_j}}
    = (1 - u)\expect{\frac{V_i}{\expect{V_i}}\wedge\frac{V_j}{\expect{V_j}}},
\end{equation*}
where the second identity is justified by the fact that $V_i$ is
bounded and $z_i\rightarrow\infty$. The proof is completed by noting
that $\prob{F_i(Z_i)>u} = 1 - u.\hfill\Box$

\subsection{Multivariate Conditional Survival Function\label{subsec:condsurv}}
For the analysis of extreme values it is of interest to calculate the multivariate conditional survival function. The following result provides the relevant expression, as a function of the angular measure.
\begin{prop}
Assume the same conditions of Proposition \ref{ppchi}. 
Let $\alpha \subset \{1, \ldots ,p\}$ be a collections of indexes. 
Then     
\begin{equation} \label{eqn:condsurv2df}
    \text{Pr}\left[\bigcap_{l \in \alpha} Z_l > z_l \right|\left. \bigcap_{l\not\in\alpha} Z_l > z_l\right] =
      \frac{\expect{\bigwedge_{k = 1}^d 1\wedge \frac{V_k}{z_k}}}{\expect{\bigwedge_{k \not\in\alpha}1\wedge\frac{V_k}{z_k}}}.
  \end{equation}
\end{prop}  
The proof uses a similar approach to the proof of Proposition \ref{ppchi}.

### 2021-10-12 methodology_pnorm.tex
update

\subsection{Projection onto an arbitrary unit hypersphere}

@@

Consider a vector $\bm{s} \in {\mathbb R}^d$, and let the $\mathcal{L}_p$-norm be defined as
  \begin{equation*}
    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
  \end{equation*}
for $p>0$. The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively. 
The $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
  \begin{equation*}
    \lVert \bm{s} \rVert_{\infty}
      = \lim\limits_{p\to\infty} \lVert \bm{s} \rVert_p
      = \max_{\ell\in\lbrace1,\ldots,d\rbrace}s_{\ell}.
  \end{equation*}
To define an angular measure,  we are interested in the direction of
vectors in ${\mathbb R}_{+}^d$, thus, we project them onto
  \begin{equation*}
    \mathcal{S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
  \end{equation*}
  the $\mathcal{L}_p$-norm unit hypersphere.
  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto $\mathcal{S}_{p}^{d-1}$ is given as
  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in \mathcal{S}_{p}^{d-1}$.
  Observations on one hypersphere can be projected without loss of information onto another by
  dividing by the defining norm of the target hypersphere. In fact, letting 
  \begin{equation*}
    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}.
  \end{equation*}
  the transformation
  \begin{equation}
    \label{eqn:pnormt}
    T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
  \end{equation}
  is invertible with
  \begin{equation}
    \label{eqn:pnormtinv}
    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right) =
      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
  \end{equation}
  The Jacobian of the transformation takes the form
  \begin{equation}
    \label{eqn:pnormjac}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
        {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on $\mathcal{S}_{p}^{d-1}$, for any finite $p>0$. Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable. Thus, even though the projected distribution on $\mathcal{S}_\infty^{d-1}$ is defined, there is no corresponding density.

>>>>

\subsection{Projection onto an arbitrary unit hypersphere\label{subsec:projection}}

@@

Consider a vector $\bm{s} \in {\mathbb R}^d$, and let the $\mathcal{L}_p$-norm be defined as
  \begin{equation*}
    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
  \end{equation*}
for $p>0$. The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively. 
The $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
  \begin{equation*}
    \lVert \bm{s} \rVert_{\infty}
      = \lim\limits_{p\to\infty} \lVert \bm{s} \rVert_p
      = \max_{\ell\in\lbrace1,\ldots,d\rbrace}s_{\ell}.
  \end{equation*}
To define an angular measure,  we are interested in the direction of
vectors in ${\mathbb R}_{+}^d$, thus, we project them onto
  \begin{equation*}
    {\mathbb S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
  \end{equation*}
  the $\mathcal{L}_p$-norm unit hypersphere.
  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto ${\mathbb S}_{p}^{d-1}$ is given as
  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in {\mathbb S}_{p}^{d-1}$.
  Observations on one hypersphere can be projected without loss of information onto another by
  dividing by the defining norm of the target hypersphere. In fact, letting 
  \begin{equation*}
    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}.
  \end{equation*}
  the transformation
  \begin{equation}
    \label{eqn:pnormt}
    T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
  \end{equation}
  is invertible with
  \begin{equation}
    \label{eqn:pnormtinv}
    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right) =
      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
  \end{equation}
  The Jacobian of the transformation takes the form
  \begin{equation}
    \label{eqn:pnormjac}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
        {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$. Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable. Thus, even though the projected distribution on ${\mathbb S}_\infty^{d-1}$ is defined, there is no corresponding density.
